---
title: "STAT 400 Notes Lecture 6"
author: Poisson Regression Part 1
output: html_document
date: "Jan 31, 2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Front Matter - Clean Environment, Load Libraries, User Defined Functions
```{r}
rm(list = ls())
library(ggplot2)
library(dplyr)
library(kableExtra) #For Example 2b

#Read in Data from Roback's github
fHH1 <- read.csv(file = "https://raw.githubusercontent.com/proback/BeyondMLR/master/data/fHH1.csv")
```

## Example 2a
```{r}

```

## Example 2b - Basic Exploration
```{r}
#Univariate Exploration
#total - summary stats
fHH1 %>% summarize(totalmean = mean(total),
                   totalvar = var(total),
                   totalsd = sd(total))

#plot of total
ggplot(data = fHH1, mapping = aes(total)) + 
  geom_histogram(binwidth = .25, color = "black", 
                 fill = "white") + 
  labs(x = "Number in the house excluding head of household",
       y = "Count of households")

#summary stats for age 
fHH1 %>% summarize(agemean = mean(age),
                   agevar = var(age),
                   agesd = sd(age))

#Plot for age
ggplot(data = fHH1, mapping = aes(x=age)) +
  geom_histogram(binwidth = 5, fill = "white", color = "black") +
  labs(x = "Age for Head of Household")

#summary statistics (counts and proprotions) for location
#Find counts
fHH1 %>%
  group_by(location) %>%
  summarize(N = n())

#Find proportions (could also use prop.table command)
fHH1 %>%
  group_by(location) %>%
  summarize(N = n()) %>%
  mutate(Proportion = N/sum(N))

#Plot locations
ggplot(data = fHH1, mapping = aes(x = location)) +
  geom_bar() +
  labs(x = "Location",
       y = "Number of Households")


#summary statistics (counts and proprotions) for roof

#Find proportions (could also use prop.table command)
fHH1 %>%
  group_by(roof) %>%
  summarize(N = n()) %>%
  mutate(Proportion = N/sum(N))

#Plot roofs
ggplot(data = fHH1, mapping = aes(x = roof)) +
  geom_bar() +
  labs(x = "Type of Roof",
       y = "Number of Households")

#Multivariate
#total by location
fHH1 %>% group_by(location) %>%
         summarize(totalmean = mean(total),
                   totalvar = var(total),
                   totalsd = sd(total))

#Plot of total by location
ggplot(data=fHH1, mapping = aes(x=location, y=total)) +
  geom_boxplot()

#total by roof
fHH1 %>% group_by(roof) %>%
         summarize(totalmean = mean(total),
                   totalvar = var(total),
                   totalsd = sd(total))

#Plot of total by roof category
ggplot(data=fHH1, mapping = aes(x=roof, y=total)) +
  geom_boxplot()
```


## Example 2b - Discretizing Age/Examining Distribution by Age Group
```{r}
cuts = cut(fHH1$age,
           breaks=c(15,20,25,30,35,40,45,50,55,60,65,70))
ageGrps <- data.frame(cuts,fHH1)
ggplot(data = ageGrps, aes(x = total)) +
  geom_histogram(binwidth = .25, color = "black", 
                 fill = "white") +
  facet_wrap(cuts) +
  xlab("Household size")
```

## Example 2b - Mean and variance by age group
```{r}
# Mean = Variance
table1chp4<- ageGrps  %>% group_by(cuts)  %>% 
  summarise(mnNum= mean(total),varNum=var(total),n=n())
kable(table1chp4, booktabs=T, 
      caption="Compare mean and variance of household size within each age group.",
      col.names = c("Age Groups", "Mean", "Variance", "n")) %>%
  kable_styling(full_width = F)
```

## Example 2c - Checking Linearity
```{r}
## Checking linearity assumption: Empirical log of the means plot
sumStats <- fHH1 %>% group_by(age) %>% 
  summarise(mntotal = mean(total),
            logmntotal = log(mntotal), n=n())
ggplot(sumStats, aes(x=age, y=logmntotal)) +
  geom_point()+
  geom_smooth(method = "loess", size = 1.5)+
  xlab("Age of head of the household") +
  ylab("Log of the empirical mean number in the house") 
```

## Example 2c - Checking linearity by location - not in notes
```{r}
sumStats2 <- fHH1 %>% group_by(age, location) %>% 
  summarise(mntotal = mean(total),
            logmntotal = log(mntotal), n=n())
ggplot(sumStats2, aes(x=age, y=logmntotal, color=location,
                      linetype = location, shape = location)) +
  geom_point()+
  geom_smooth(method = "loess", se=FALSE)+
  xlab("Age of head of the household") +
  ylab("Log empirical mean household size") 
```


## Example 2d - Build Poisson Model using Age
```{r}
model1 = glm(total ~ age, family = poisson, data = fHH1)
summary(model1)
```

## Example 2f 

```{r}
confint(model1, level = 0.95) #for coefficient of age
```

## Example 2g

```{r}
exp(confint(model1, level = 0.95)) #for relative risk
```

## Example 2h - Prediction for age=65
```{r}

```

