---
title: "STAT 400 Notes Lecture 5"
author: Likelihoods Part 2
output: html_document
date: "Jan 29, 2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter - Clean Environment, Load Libraries, User Defined Functions
```{r}
rm(list = ls())
library(tidyverse)

#Read in Real Estate data
real.estate <- read.table(file="https://online.stat.psu.edu/onlinecourses/sites/stat501/files/data/realestate.txt",header=TRUE)
```

## Example 2a
```{r}
#Create New Variables, including interaction
real.estate <- real.estate %>%
                    mutate(log10Price = log10(SalePrice),
                           log10SqFeet = log10(SqFeet),
                           log10FeetPool = log10SqFeet*Pool)

#Build Models
model1 <- lm(log10Price ~ log10SqFeet + Pool + log10FeetPool, data=real.estate)
model2 <- lm(log10Price ~ log10SqFeet, data=real.estate)

#Compare Models
anova(model2, model1)
```

1. Hypotheses
    - Null Hypothesis: $H_0: \beta_2 = \beta_3 = 0$
    - Alternative Hypothesis: $H_1: \beta_j \ne 0 \text{ for at least 1 } j \in \{2,3\}$

\

2. Test Statistic: $t=2.7086$

3. p-value: $0.06757$

4. Decision: Depends on $\alpha$. Using $\alpha = 0.05$, we fail to reject $H_0$. 

5. Conclusion: There is not enough evidence to reject that $\beta_2 = \beta_3 = 0$. In other words, there is not enough evidence to reject that the the reduced model, which contains only the log base 10 of square feet, is sufficient.

## Example 3a
```{r}
#Find the log likelihood for the reduced model (model2)
logLik(model2)
```

## Example 3b
```{r}
#Find the log likelihood for the full model (model1)
logLik(model1)
```


## Example 4
```{r}
#Compute LRT Test stat
LRT_stat <- -2*(as.numeric(logLik(model2))-as.numeric(logLik(model1)))
LRT_stat

#Compute LRT p-value
1 - pchisq(q=LRT_stat,df=2)
```


## Example 6
```{r}
#Compare models 1 and 2 using AIC
AIC(model1) #Full
AIC(model2) #Reduced

#Compare models 1 and 2 using BIC
BIC(model1) #Full
BIC(model2) #Reduced
```

