---
title: "STAT 400 Notes Lecture 3"
author: Regression Review Part 3
output: html_document
date: "Jan, 2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter - Clean Environment, Load Libraries, User Defined Functions
Remember to install package first:
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")

```{r}
rm(list = ls()) #CAUTION: command erases everything in environment

#Put all library commands here
library(tidyverse)

#Read in all files here 
derby <- read.csv(file = "https://raw.githubusercontent.com/proback/BeyondMLR/master/data/derbyplus.csv")
```

## Code from Lecture 2
```{r}
#Create New Variables
derby <- derby %>% 
            mutate(fast = ifelse(condition == "fast", 1, 0),
                   good = ifelse(condition == "good", 1, 0),
                   fastfactor = ifelse(condition == "fast", "fast", "not fast"),
                   yearnew = year - 1896)
```

## Example 1a - How are the winning speeds distributed?
```{r}
ggplot(data = derby, 
       mapping = aes(x = speed)) +
  geom_histogram(color = "black", fill = "lightblue", binwidth = 0.5) +
  labs(x = "Winning Speed",
       y = "Number of Races")

```

The distribution of winning speeds appears to be left skewed. The majority of winning speeds are between 51 and 55 ft/s. 

The mean winning speed is `r mean(derby$speed)`. The standard deviation of winning speeds is `r sd(derby$speed)`.

## Example 1b - Are the winning speeds affected by the track condition? 
```{r}
ggplot(data = derby, mapping = aes(x = condition, y = speed)) +
  geom_boxplot() +
  labs(x = "Track Condition",
       y = "Winning Speed in ft/s")

```


```{r}
#Calculate summary statistics for each track condition
derby %>%
  group_by(condition) %>%
  summarize(N = n(),
            MeanSpeed = mean(speed, na.rm = TRUE),
            MedSpeed = median(speed, na.rm = TRUE),
            sdSpeed = sd(speed, na.rm = TRUE))
```


## Example 1c - What is the relationship between winning speed, year, and whether track conditions were fast?
```{r}
#Create coded scatterplot
ggplot(data = derby, 
       mapping = aes(x = year, y = speed, 
                     color = fastfactor,
                     shape = fastfactor)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Year",
       y = "Winning Speed in ft/s",
       color = "Condition",
       shape = "Condition")

```

```{r, eval = FALSE}
ggplot(data = derby, 
       mapping = aes(x = year, y = speed, 
                     color = as.factor(fast),
                     shape = as.factor(fast))) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Year",
       y = "Winning Speed in ft/s",
       color = "Condition",
       shape = "Condition")
```


Informal Observations:

- red line is above blue. both slopes are positive. blue line has a steeper slope

Formal Observations:

- We expect faster winning times for fast track conditions. 

- Under all track conditions, the the winning speeds are increasing over time. 

- The rate of change of the speed with respect to the year (i.e., slope) is higher for good/slow conditions than fast conditions.

- Since the lines are different this would suggest that track conditions should be incorporated into the model for speed.

## Example 2a - Have winning speeds increased at a constant rate since 1896? 
Write the true/population model. Define your notation.

The model is given by:

$$y_{i} = \beta_{0} + \beta_{1} x_{i,year} + \epsilon_{i}$$


where 

$$
\begin{aligned}
y_{i} &= \text{the winning speed in ft/s for the } i^{th} \text{year} \\
x_{i,year} &= \text{year of the } i^{th} \text{race}
\end{aligned}
$$

## Example 2b - Build Linear Regression Model
```{r}
#Build model
model1 <- lm(speed ~ year, data=derby)

#Summarize results
summary(model1)
```

The estimated regression equation is given by:

$$\hat{y}_{i} = 2.05 + 0.026x_{i}$$

## Example 2e - Is there a statistically significant relationship between the year and the winning speed?

Interpret the hypothesis test associated with the slope.

1. Hypotheses
    - Null Hypothesis: $H_0: \beta_1 = 0$
    - Alternative Hypothesis: $H_1: \beta_1 \ne 0$

\

2. Test Statistic: $t=11.251$

3. p-value: $p < 2.2 \times e^{-16}$

4. Decision: Reject $H_0$. 

5. Conclusion: There is sufficient evidence to reject that the intercept only model is sufficient. We should use the model that includes the year.

## Example 2f - Confidence Interval for slope
```{r}
confint(model1, level = 0.95)
```

## Example 2i - What does recentering the year change?
```{r}
#Build model
model1a <- lm(speed ~ yearnew, data=derby)

#Summarize results
summary(model1a)
```

## Example 2j - Inspect the assumptions
```{r}
par(mfrow=c(2,2)) #Tells R to display next 4 plots in 2x2 grid
plot(model1, which = c(1, 2, 4))
par(mfrow = c(1,1)) #Tells R to display plots in a "1x1" grid
```


## Example 3a - Building Models
```{r}
#Model1
model1recentered <- lm(speed ~ yearnew, data=derby)

#Model2
model2 <- lm(speed ~ fast, data = derby)

#Model3
model3 <- lm(speed ~ yearnew + fast, data = derby)

derby <- 
  derby %>%
  mutate(yearnewFast = yearnew*fast)
#Model4
model4 <- lm(speed ~ yearnew + fast + yearnewFast, data = derby)
```




## Example 3b - Comparing Models 3 and 4
```{r}
#Compare summaries
summary(model3)
summary(model4)

#Significance Test
anova(model3, model4)
```


## Example 3c - Confidence Intervals for Model 3
```{r, eval = FALSE}
#Find Confidence Interval
confint(model3)
```













